---
title: "Notes on Discrete Choice with Capacity Constraints"
author: "Nick Mader, Chapin Hall at the University of Chicago (nmader@chapinhall.org)"
date: "April 6, 2017"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(data.table)
library(magrittr)
library(broom)

# A LaTeX reference for math symbols:
# http://web.ift.uib.no/Teori/KURS/WRK/TeX/symALL.html

# To dos --
# 2. rerun analysis with updated assignment rule
# 3. create new data structure for estimation with many possibilities

```

# Statement of the Problem

In policy problems around takeup of human services, capacity constraints of service providers may lead to underestimation of the full demand for services, if households do not take up services if constrained out of enrollment. In this brief, I consider the example of family enrollment of children in Head Start ("HS") programming.

Let $i$ generically index children to be enrolled in programming, with a total number of $N$ children. Children and families have traits $x_i$, and their constrained choice to enroll in HS is

$$
y_i = 1 \Longleftrightarrow y_i^* \geq 0 \\
y_i^* = x_i^{\prime}\beta + \epsilon_i
$$

Where the parameter vector $\beta$ represents weights translating observed characteristics into latent propensity to seek enrollment in services.

However, we now impose a capacity constraint where

$$\sum_{j=1}^N{y_i} \leq \bar{n}$$

. Thus, an acceptance rule must be imposed, where by families with $y_i^*$ are allowed to enroll. This rule for assigning probability of enrollment for a given family $r_i$ is dependent on other families with $y_i^* \geq 0$:

$$
\begin{aligned}
r_i(y_i^* \| \{\{y_j^*\} \| y_j^* \geq 0 \forall j=1,\dots,N \}) =
\begin{matrix}
f(y_{j\neq i}^*), f\in(0,1) & if & y_i^* \geq 0
0 & if & y_i^* < 0 \\
\end{matrix}
\end{aligned}
$$

where the full 

# Single Provider with Random Acceptance

We develop a sample economy with $N=100$ agents, a single provider, provider capacity of $\bar{n}=25$, and assigment rule of

$$
r_i(\textbf{y}_j^*, \forall \{j|y_j^* \geq 0 \})=
\begin{matrix}
0 & if & y_i^*<0 \\
\bar{n}/\tilde{n} ,\; \tilde{n} = max(\Sigma_j\{y_j^* \geq 0\}, \; \bar{n})  & if & y_i^* \geq 0 
\end{matrix}
$$

to demonstate the bias of conventional methods. 

```{r simple model setup}
set.seed(60607)
N <- 1000
nbar <- 250
b <- 5
a <- -2

RunLottery <- function(ystars, nbar){
  ystars_ge0 <- 1*(ystars > 0)
  nInterested <- sum(ystars_ge0)
  prob <- nbar / nInterested
  
  draw <- runif(length(ystars_ge0))
  draw <- ifelse(ystars_ge0 == 0, 1, draw)
  enr <- ifelse(ystars_ge0 == 0, 0, 1*(draw <= sort(draw)[nbar]))
  return(enr)
}

MakeData <- function(){
  
  # Draw basic features
  x <- runif(N)
  e <- rlogis(N)
  ystar = a + x*b + e
  ystar_ge0 = 1*(ystar >= 0)
  
  # Generate data set
  dt <- 
    data.table(x = x,
               e = e,
               ystar = ystar,
               ystar_ge0 = ystar_ge0)
    
  # Add capacity contraint
  dt$y <- RunLottery(ystars = dt$ystar, nbar = nbar)
  
  
  return(dt)
}

#summary(dt$ystar_ge0)
#sum(dt$y)

dt <- MakeData()

dt[, summary(ystar_ge0)]

```

<!-- Visualize enrollments -->
```{r simple model data visualization}
ggplot(dt, aes(x = x, y = ystar)) +
  geom_point(aes(size = factor(ystar_ge0), colour = factor(y))) +
  geom_hline(yintercept = 0) +
  geom_smooth(method = "loess") +
  scale_size_manual(name = "", values = c(1, 2), labels = c("Not Interested", "Interested")) +
  scale_colour_manual(name = "", values = c("black", "blue"), labels = c("Not Enrolled", "Enrolled"))

```


```{r simple model inference}
simpleReg <- glm(y ~ x, data = dt, family = "binomial")
tidy(simpleReg)
```

<!-- /!\ Demonstrate fix. Basically--modify the individual likelihoods -->

The proper likelihood function is

$$
\begin{aligned}
\mathcal{L} &= \prod_i^N\Big[ pr(y_i^*<0)+pr(y_i^*>0)\Big(1-\bar{n}/\tilde{n}\Big)\Big]^{y_i=0}
  \Big[pr(y_i^*>0)\Big(\bar{n}/\tilde{n}\Big)\Big]^{y_i=1} \\
\tilde{n} &= \sum_i^N (y_i^*>0)
\end{aligned}
$$

```{r simple model fix}

fullLike <- function(beta, dt){
  xb <- beta[1] + dt$x*beta[2]
  pr_ysGe0 <- plogis(xb)
  r <- min(1, nbar/sum(pr_ysGe0)) # probability of being accepted
  
  pr_yLt0 <- (1-pr_ysGe0) + pr_ysGe0*(1-r)
  pr_yGe0 <- pr_ysGe0*r
  
  ll <- (1-dt$y)*log(pr_yLt0) + dt$y*log(pr_yGe0)
  return(-sum(ll))
}

# Run a range of data draw scenarios to run estimation with

runs <- 100
m <- matrix(rep(0, 4*runs), ncol = 2)
mypar <- data.frame(alpha = m[, 1], beta = m[, 2], stringsAsFactors = FALSE)
for (i in 1:100){
  if (i %% 10 == 0) print(paste("Working on run", i))
  dt <- MakeData()
  
  # Run naive
  logistic <- glm(y ~ x, data = dt, family = "binomial")
  mypar[2*i-1, ] <- c(logistic$coefficients)
  
  # Run capacity-aware estimation
  opt <- optim(c(0, 0), fullLike, dt = dt)
  mypar[2*i,] <- c(opt$par)
}
mypar$spec <- c("naive", "adj")

true <- data.frame(alpha = a, beta = b, spec = "true", stringsAsFactors = FALSE)
mypar <- rbind(mypar, true) %>%
  within(spec <- factor(spec, levels = c("naive", "adj", "true")))

# Display results

# /!\ Look for how to change the order of 
ggplot(data = mypar, aes(x = alpha, y = beta, color = factor(spec), size = factor(spec))) +
  geom_point() +
  scale_size_manual(values = c(1.5, 1.5, 4), guide = FALSE) +
  scale_color_discrete(name = "", labels = c("Naive", "Adjusted", "True")) +
  labs(x = "alpha", y = "beta") +
  ggtitle("Comparison of Monte Carlo Model Results vs True Parameters")

```


# Extension to Non-Random Admission

<!-- Basically, whoever wants it more, as something like probability being proportional to max(y*,0) -->

# Extension to Multiple Service Providers

Now let $j$ generically index choices, which number $J$.

```{r multiple providers data setup}

# Simulated data creation for the DSSG school choice project is at -- https://github.com/dssg/predicting_student_enrollment/blob/e7938399ffc67fddc7e7fc62d11ed757f65fa6c9/estimation-sandbox/gen-sim-data-for-estimation.R

J <- 10
nbar_j <- round(50*runif(J) + 1, 0)
#table(nbar_j)

RunMultiLottery <- function(vs, nbar_j){
  # Plan is:
  # (1) identify first choices for everyone. 
  # (2) run lotteries for each case where there's a capacity contraint to determine who gets in
  # (3) move households to their second choice, if they have them
  # (4) if second choices aren't already all full, then assign and/or run lotteries
  # (5) run steps 2-4 until either everyone is assigned or no household has a remaining choice of interest
}

MakeMultiData <- function(){
  
  # Draw basic features
  x <- runif(N)
  z <- runif(J)
  
  # Generate basic data
  dt <- merge(x, z) %>% # With no "by", this generates a permutation set
    rename(z = y) %>%
    within({
      xz <- x*z
      d <- 10*runif(length(x))
      e <- rlogis(length(x))
      v_ij <- -3 + x + 2*z + 1.5*xz - 0.1*d + e
    })
  summary(dt$v_ij)
  mean(dt$v_ij > 0)
  
  # Establish desired and actual enrollments
  
  
  # Add capacity contraint
  dt$y <- RunLottery(ystars = dt$ystar, nbar = nbar)
  
  return(dt)
}

```


